---
title: "Golden Mussel Habitat Suitability Assessment South America"
author: "Chris Madsen & John Phelan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)


library(rinat)
library(tidyverse)
# library(DBI)
library(bcinvadeR)
library(terra)
library(sf)
library(geodata)
library(ggpubr)
library(ecospat)
library(ENMeval)
library(readxl)
library(ENMTools)
library(data.table)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggthemes)
library(kableExtra)
library(modEvA)
library(rSDM)
source("./scripts/utils/select_min_max_temp_rast_f.R")
source("./scripts/utils/select_TMax_min_temp_rast_f.R")
# Set up file paths
lan_root = "//SFP.IDIR.BCGOV/S140/S40203/RSD_ FISH & AQUATIC HABITAT BRANCH/General/"
proj_wd = getwd()
onedrive_wd = paste0(str_extract(getwd(),"C:/Users/[A-Z]+/"),"OneDrive - Government of BC/data/")

# Options
reclip_hydroclim = FALSE

reclip_slope = FALSE

reclip_soil = FALSE


rerun_maxent_in_south_america = TRUE

reclip_min_temp = FALSE

# Vectors of country (and province) names
asian_countries = c("China","Hong Kong","Japan","South Korea","Laos","Thailand","Taiwan")

south_american_countries = c("Argentina","Brazil","Paraguay","Uruguay")

north_america = c("United States","BC")

chosen_rasters = c("slope_average","soil_pH_in_H2O")

other_rasters<-c("max_temp_warmest_month", "clay_content_mass_fraction", "silt_content_mass_fraction", "sand_content_mass_fraction","soil_organic_carbon", "soil_organic_carbon")

source("scripts/utils/get_occurrence_records.R")

source("scripts/utils/clip_predictor_data_to_areas_of_interest.R")

#source("scripts/utils/read_in_predictor_data.R")

source("scripts/utils/split_occurrence_records_by_area.R")

source("scripts/utils/read_in_predictor_data_asia.R")

source("scripts/utils/get_pseudo_occurrences.R")

#asian_country_r$annual_mean_temp <- asian_country_r$annual_mean_temp/10
#asian_country_r$max_temp_warmest_month <- asian_country_r$max_temp_warmest_month/10


invisible(gc())



# ggplot()+
#   geom_map(data = world_map, map= world_map, aes(long, lat, map_id = region))

```



## Modelling South America occurrences

```{r get_south_preds, include=FALSE}

invisible(source("scripts/utils/read_in_predictor_data_south_america.R"))

south_america_temps<-select_temp_rast(south_american_countries)
south_america_r<-c(south_america_r, south_america_temps)

SA_max_temps<-select_max_min_temp_rast(south_american_countries)
names(SA_max_temps) <- "Max_of_max_temp"
south_america_r<-c(south_america_r, SA_max_temps)
rm(SA_max_temps)


```

```{r get_sa_points, include = FALSE}


sa_occs_sf<- st_as_sf(x = south_american_occs, coords = c("x","y"), crs = 4326)
sa_occs_sf<-st_transform(sa_occs_sf, crs(south_america_r))
sa_occs_moved<-points2nearestcell(sa_occs_sf, south_america_r, layer = 1, move = TRUE, distance = 2000)

occmap(sa_occs_sf,south_america_r[[1]], pcol = "black")
occmap(sa_occs_moved, add = TRUE, pcol = "red")

sa_occs_mv<- sa_occs_moved |> 
  mutate(x = sf::st_coordinates(geometry)[,1], y = sf::st_coordinates(geometry)[,2]) |> 
  st_drop_geometry()

pointDiff<-setdiff(sa_occs_mv, south_american_occs)


```


```{r maxent_South_America}



if(rerun_maxent_in_south_america){
  
  sa_pseudos = predicts::backgroundSample(south_america_r,
                                     p = terra::vect(south_american_occs, geom = c("x", "y")),
                                     n = 10000,
                                     extf = 0.9,
                                     tryf = 500) |>
  tidyr::as_tibble()
  
  sa_model = ENMevaluate(occs = sa_occs_mv,
                           envs = south_america_r,
                           bg = sa_pseudos,
                           algorithm = 'maxent.jar',
                           partitions = 'block',
                           #parallel = TRUE,
                           tune.args = list(fc = c('L','LQ'),
                                            rm = c(1:5)))
  saveRDS(sa_model, "Golden_Mussel_MaxEnt_trained_on_SA_Occurrences.rds")
  invisible(gc())
  # Chris' poor machine can barely run the above maxent model.
  # it took a bit of time for it to run on Jonh's computer too.can i run it
  #It was not at all fast.
} else {
  sa_model = readRDS("Golden_Mussel_MaxEnt_trained_on_SA_Occurrences.rds")
}
invisible(gc())



```

```{r maxent_sa_vars}

# The resulting 5 models seem to all have the same auc.train. Let's just use the first one.
#asia_best_model <- asia_model@models[[which.min(asia_model@results$AICc)]]



sa_auc_cbi <- sa_model@results|> 
  mutate(auc_cbi = (as.numeric(auc.train) + as.numeric(cbi.train) )/ 2) |> 
  arrange(desc(auc_cbi)) 

topfc<-as.character(sa_auc_cbi[1,]$fc)
  toprm<-as.character(sa_auc_cbi[1,]$rm)
  
sa_best_model <- sa_model@models[[which(sa_model@results$fc == topfc & sa_model@results$rm == toprm)]]
 
sa_html<-sa_best_model@html
invisible(file.copy(from = sa_html, to = paste0("./output/sa_MaxEnt_results.html")))

model_results_df = as.data.frame(sa_best_model@results)
model_results_df$var = row.names(model_results_df)
sa_model_AUC = model_results_df[model_results_df$var == 'Training.AUC',]$V1

sa_best_model_preds = terra::rast(sa_model@predictions[[which.min(sa_model@results$AICc)]])


```

### Model performane {.tabset}

#### Model settings

Note: Model parameters include:

Tuning Settings:
```{r sa_tune}
sa_model@tune.settings |> kbl() |> kable_minimal() 

```


#### Summary details
Results:
```{r sa_summary}

summary_models <- sa_model@results |>  dplyr::select(fc, rm, auc.train, cbi.train, AICc, delta.AICc, w.AIC, ncoef) |> data.frame()
# asia_model@results |>  dplyr::select(fc, rm, auc.train, cbi.train, AICc, delta.AICc, w.AIC, ncoef) |> 
selected_row <- which(summary_models$fc == topfc & summary_models$rm == toprm)

summary_models |> 
  kableExtra::kbl() |> 
  kableExtra::kable_minimal() |> 
  kableExtra::row_spec(selected_row, bold = TRUE) |> 
  pack_rows("Selected model", selected_row, selected_row, label_row_css = "background-color: #666; color: #fff;") |> 
  column_spec(c(4, 5), background = "lightgrey")

  
```

#### Full report

```{r sa_full}
sa_model@results |> kbl() |> kable_minimal() |> kableExtra::row_spec(selected_row, bold = TRUE) |> 
  pack_rows("Selected model", selected_row, selected_row, label_row_css = "background-color: #666; color: #fff;") |> 
  column_spec(c(4,5), background = "lightgrey")
```

#### Variables contribution 

Variable Importance in top model:
```{r sa_var_contribute}
sa_best_model_var <- sa_model@variable.importance[[which(sa_model@results$fc == topfc & sa_model@results$rm == toprm)]]

sa_best_model_var <- sa_best_model_var |> 
  mutate(variable = case_when(
    variable == "Clay_content_mass_fraction" ~ "clay_content",
    variable == "Min.monthlyairtemperatureforJanuary" ~ "min_air_temp_mins",
    variable == "Min.monthlyairtemperatureforJuly" ~ "Max_air_temp_mins",
    variable == "Soil_organic_carbon" ~ "soil_organic_carbon",
    variable == "Soil_P_h_in_H2O" ~ "soil_ph",
    variable == "annual_mean_temp" ~ "annual_mean_temp",
    variable == "slope_average" ~ "slope_avg",
    TRUE ~ variable
  ))
  
sa_best_model_var |> dplyr::arrange(desc(percent.contribution)) |> kbl() |> kable_minimal()

```

#### Variable response curves
```{r sa_repsonse_curve}

response(sa_model@models[[which(sa_model@results$fc == topfc & sa_model@results$rm == toprm)]])


```


## Model Results - South America

The model has been trained on:

* Annual mean temperature
* Average slope
* pH
* Minimum temperature by month by country (coldest month, coldest average temperature)
* warmest average temperature of warmest month by country (warmest month, warmest average temperature)


I haven't tested the colinearity of temperature as of yet, so this should be done and the temperature rasters changed if required.

### Training Model on South America Occurrence Data {.tabset}



#### MaxEnt Model for South America Extent

```{r maxent_plot_sa}

sa_bound<-ne_countries(continent = "south america", returnclass = "sf", scale = "medium")
rast_ext<-ext(sa_best_model_preds)
map_ext<-ext(sa_bound)



ggplot() +
  geom_sf(data = sa_bound, fill = "white", color = "grey", size = 0.1) +  # Keep land white
  tidyterra::geom_spatraster(data = sa_best_model_preds) +
  geom_point(data = south_american_occs, aes(x = x, y = y), size = 2, alpha = 0.5) +
  coord_sf(xlim = c(xmin(rast_ext), xmax(rast_ext)), 
           ylim = c(ymin(rast_ext), ymax(rast_ext))) +
  scale_fill_viridis_c(option = 'turbo', limits = c(0.01, 1), na.value = "transparent") +
  labs(fill = 'Suitable\nHabitat') +
  theme_map()+
  theme(plot.background = element_rect(fill = 'lightblue'))

```


```{r thresholds}
# Identify what the minimum predicted suitability was for the known occurrence points.
#asian_min_ten_percent_thresh = invisible(min(terra::extract(asia_best_model_preds, asian_occs), na.rm=T))

result_sa<-sa_best_model@results |> as.matrix() |> as.data.frame()
result_sa = result_sa |> 
    dplyr::mutate(metric = snakecase::to_snake_case(rownames(result_sa))) |>
    dplyr::rename(value = V1) |>
    tidyr::as_tibble() |>
    dplyr::select(metric, value) 


hab_thres<- result_sa |> dplyr::filter(metric %in% "x_10_percentile_training_presence_cloglog_threshold")
min_thresh <- result_sa |> dplyr::filter(metric %in% 
                                             "minimum_training_presence_cloglog_threshold")


#sa_min_ten_percent_thresh <- hab_thres$value
sa_min_ten_percent_thresh <-min_thresh$value



```


#### MaxEnt Model Transformed: "Habitat-or-Not" Threshold Map
```{r maxent_asia_habitat}
# Apply to convert the predictions for Asia into habitat/not habitat
sa_preds_hab_or_not = sa_best_model_preds > sa_min_ten_percent_thresh

# asia_min_hab = asia_best_model_preds > asian_min_occ_threshold

sa_preds_hab_or_not <- terra::as.factor(sa_preds_hab_or_not)
levels(sa_preds_hab_or_not) <- list(data.frame(ID = c(0, 1), fc.LQ_rm.1 = c("FALSE", "TRUE")))


ggplot() +
  geom_sf(data = sa_bound, fill = "white", color = "grey", size = 0.1) +  # Land boundary
  geom_point(data = sa_occs_mv, aes(x = x, y = y), size = 2, alpha = 0.5) +  # Observation points
  tidyterra::geom_spatraster(data = sa_preds_hab_or_not) +  # Raster layer
  coord_sf(xlim = c(xmin(rast_ext), xmax(rast_ext)), 
           ylim = c(ymin(rast_ext), ymax(rast_ext))) +
  scale_fill_manual(
    values = c("FALSE" = "lightgreen", "TRUE" = "purple"),  # Discrete color scale
    na.value = "transparent"  # Handle NA values
  ) +
  labs(fill = 'Suitable\nHabitat') +
  theme_map()+
  theme(plot.background = element_rect(fill = 'lightblue'))

```

#### MaxEnt Model for North American Extent

```{r get_north_preds, include=FALSE}
source("scripts/utils/read_in_predictor_data_north_america.R")

north_america_temps<-select_temp_rast(north_america)

north_america_r<-c(north_america_r, north_america_temps)

NA_max_temps<-select_max_min_temp_rast("united states")
names(NA_max_temps) <- "Max_of_max_temp"
north_america_r<-c(north_america_r, NA_max_temps)
rm(NA_max_temps)

```


```{r predict_north_america_and_plot}
# ---------------------------------------
# STEP 3: APPLY MODEL TO NORTH AMERICA
# ---------------------------------------

# Apply the Asia model to North America.
north_america_preds = terra::predict(north_america_r, sa_best_model, type = "response", na.rm=T)

north_amer_map<-ne_countries(continent = "north america", returnclass = "sf", scale = "medium")
rast_ext<-ext(north_america_preds)
map_ext<-ext(north_amer_map)

ggplot()+
  geom_sf(data = north_amer_map, fill = "white", color = "grey", size = 0.1) +
  tidyterra::geom_spatraster(data = north_america_preds) +
  geom_point(data = north_american_occs, aes(x = x, y = y), size = 2, alpha = 0.5) +
  coord_sf(xlim = c(xmin(rast_ext), xmax(rast_ext)), 
           ylim = c(ymin(rast_ext), ymax(rast_ext))) +
  scale_fill_viridis_c(option = 'turbo', limits = c(0.01, 1), na.value = "transparent") +
  labs(fill = 'Suitable\nHabitat') +
  theme_map()+
  theme(legend.position = "right",
        plot.background = element_rect(fill = 'lightblue'))


```

#### MaxEnt Model Transformed: "Habitat-or-Not" Threshold Map

```{r habitat_north_america}
# Apply the minimum suitable habitat for which we found occurrences in asia,
# to calculate a habitat-or-not for north America.

north_america_preds_hab_or_not = invisible(north_america_preds > sa_min_ten_percent_thresh)

north_america_preds_hab_or_not <- terra::as.factor(north_america_preds_hab_or_not)
levels(north_america_preds_hab_or_not) <- list(data.frame(ID = c(0, 1), fc.LQ_rm.1 = c("FALSE", "TRUE")))


ggplot() +
  geom_sf(data = north_amer_map, fill = "white", color = "grey", size = 0.1) +  # Land boundary
  geom_point(data = south_american_occs, aes(x = x, y = y), size = 2, alpha = 0.5) +  # Observation points
  tidyterra::geom_spatraster(data = north_america_preds_hab_or_not) +  # Raster layer
  coord_sf(xlim = c(xmin(rast_ext), xmax(rast_ext)), 
           ylim = c(ymin(rast_ext), ymax(rast_ext))) +
  scale_fill_manual(
    values = c("FALSE" = "lightgreen", "TRUE" = "purple"),  # Discrete color scale
    na.value = "transparent"  # Handle NA values
  ) +
  labs(fill = 'Suitable\nHabitat') +
  theme_map()+
  theme(legend.position = "right",
        plot.background = element_rect(fill = 'lightblue'))

```

### Applying Trained Model on just British Columbia {.tabset}

#### MaxEnt Model for BC Extent

```{r predict_bc_and_plot}
bc_preds = terra::mask(terra::crop(north_america_preds, bc), bc)

bc_map <- bcmaps::bc_bound()
map_ext<-ext(bc_map)

ggplot() +
  geom_sf(data = bc_map, fill = "white", color = "grey", size = 0.1) +
  tidyterra::geom_spatraster(data = bc_preds) +
  coord_sf(xlim = c(xmin(map_ext), xmax(map_ext)), 
           ylim = c(ymin(map_ext), ymax(map_ext))) +
  scale_fill_viridis_c(option = 'turbo', limits = c(0.01, 1), na.value = "transparent") +
  labs(fill = 'Suitable\nHabitat') +
  theme_map()+
  theme(legend.position = "right",
        plot.background = element_rect(fill = 'lightblue'))
```

#### MaxEnt Model Transformed: "Habitat-or-Not" Threshold Map

```{r bc_habitat}
# Apply the minimum suitable habitat for which we found occurrences in asia,
# to calculate a habitat-or-not for north America.

bc_preds_hab_or_not = invisible(bc_preds > sa_min_ten_percent_thresh)

bc_preds_hab_or_not <- terra::as.factor(bc_preds_hab_or_not)
levels(bc_preds_hab_or_not) <- list(data.frame(ID = c(0, 1), fc.LQ_rm.1 = c("FALSE", "TRUE")))


ggplot()+
  geom_sf(data = bc_map, fill = "white", color = "grey", size = 0.1) +  # Land boundary
  tidyterra::geom_spatraster(data = bc_preds_hab_or_not) +  # Raster layer
  coord_sf(xlim = c(xmin(map_ext), xmax(map_ext)), 
           ylim = c(ymin(map_ext), ymax(map_ext))) +
  scale_fill_manual(
    values = c("FALSE" = "lightgreen", "TRUE" = "purple"),  # Discrete color scale
    na.value = "transparent"  # Handle NA values
  ) +
  labs(fill = 'Suitable\nHabitat') +
  theme_map()+
  theme(legend.position = "right",
        plot.background = element_rect(fill = 'lightblue'))

```
