---
title: "Golden Mussel Habitat Suitability Assessment"
author: "Chris Madsen & John Phelan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)


library(rinat)
library(tidyverse)
# library(DBI)
library(bcinvadeR)
library(terra)
library(sf)
library(geodata)
library(ggpubr)
library(ecospat)
library(ENMeval)
library(readxl)
library(ENMTools)
library(data.table)

# Set up file paths
lan_root = "//SFP.IDIR.BCGOV/S140/S40203/RSD_ FISH & AQUATIC HABITAT BRANCH/General/"
proj_wd = getwd()
onedrive_wd = paste0(str_extract(getwd(),"C:/Users/[A-Z]+/"),"OneDrive - Government of BC/data/")

# Options
reclip_hydroclim = FALSE

reclip_slope = FALSE

reclip_soil = FALSE

rerun_maxent_in_asia = TRUE

# Vectors of country (and province) names
asian_countries = c("China","Hong Kong","Japan","South Korea","Laos","Thailand","Taiwan")

south_american_countries = c("Argentina","Brazil","Paraguay","Uruguay")

north_america = c("United States","BC")

chosen_rasters = c("annual_mean_temp","max_temp_warmest_month", "slope_average",
                    "clay_content_mass_fraction", "silt_content_mass_fraction", 
                   "sand_content_mass_fraction", "soil_organic_carbon", "soil_pH_in_H2O")

source("scripts/utils/get_occurrence_records.R")

source("scripts/utils/clip_predictor_data_to_areas_of_interest.R")

#source("scripts/utils/read_in_predictor_data.R")

source("scripts/utils/split_occurrence_records_by_area.R")

source("scripts/utils/read_in_predictor_data_asia.R")

source("scripts/utils/get_pseudo_occurrences.R")

asian_country_r$annual_mean_temp <- asian_country_r$annual_mean_temp/10
asian_country_r$max_temp_warmest_month <- asian_country_r$max_temp_warmest_month/10

invisible(gc())

```

# Results {.tabset}

## Model Details 

```{r maxent_asia}

# ---------------------------------------
# STEP 1: TRAIN MODEL ON ASIAN DATA
# ---------------------------------------



if(rerun_maxent_in_asia){
  asia_model = ENMevaluate(occs = asian_occs,
                           envs = asian_country_r,
                           bg = asian_pseudos,
                           algorithm = 'maxent.jar',
                           partitions = 'block',
                           #parallel = TRUE,
                           tune.args = list(fc = c('LQ'),
                                            rm = c(1:3)))
  saveRDS(asia_model, "Golden_Mussel_MaxEnt_trained_on_Asian_Occurrences.rds")
  invisible(gc())
  # Chris' poor machine can barely run the above maxent model.
  # it took a bit of time for it to run on Jonh's computer too. It was not at all fast.
} else {
  asia_model = readRDS("Golden_Mussel_MaxEnt_trained_on_Asian_Occurrences.rds")
}
invisible(gc())

# The resulting 5 models seem to all have the same auc.train. Let's just use the first one.
asia_best_model <- asia_model@models[[which.min(asia_model@results$AICc)]]

model_results_df = as.data.frame(asia_best_model@results)
model_results_df$var = row.names(model_results_df)
asia_model_AUC = model_results_df[model_results_df$var == 'Training.AUC',]$V1

asia_best_model_preds = terra::rast(asia_model@predictions[[which.min(asia_model@results$AICc)]])
```

Note: Model parameters include:

Tuning Settings:
```{r}
asia_model@tune.settings |> knitr::kable()
```

Results:
```{r}
asia_model@results |> knitr::kable()
```

Variable Importance in top model:
```{r}
asia_model@variable.importance[[which.min(asia_model@results$AICc)]] |> knitr::kable()
```

Note: This model's Training Area-Under-Curve (AUC) value was `r asia_model_AUC` (values nearer to 1 represent models that are better able to accurately predict the occurrence points given the predictor variables, and an AUC value of 0.5 represents random predictions)

## Model Results

### Training Model on Asian Occurrence Data {.tabset}



#### MaxEnt Model for Asian Extent

```{r maxent_plot_asia}
ggplot() + 
  tidyterra::geom_spatraster(data = asia_best_model_preds) + 
  geom_point(data = asian_occs, aes(x=x, y=y)) + 
  scale_fill_distiller(palette = 'Spectral') + 
  labs(fill = 'Suitable\nHabitat')
```

```{r}
# Identify what the minimum predicted suitability was for the known occurrence points.
asia_min_suitability_with_occ = invisible(min(terra::extract(asia_best_model_preds, asian_occs), na.rm=T))
```

Note: minimum predicted suitability with a known occurrence of Golden Mussel: `r asia_min_suitability_with_occ`.

#### MaxEnt Model Transformed: "Habitat-or-Not" Threshold Map
```{r maxent_asia_habitat}
# Apply to convert the predictions for Asia into habitat/not habitat
asia_preds_hab_or_not = asia_best_model_preds > asia_min_suitability_with_occ

ggplot() + 
  tidyterra::geom_spatraster(data = asia_preds_hab_or_not) + 
  geom_point(data = asian_occs, aes(x=x, y=y)) + 
  scale_fill_manual(values = c("FALSE" = "lightgreen", "TRUE" = "purple"), na.value = 'transparent') +
  labs(fill = 'Suitable\nHabitat')
```

### Applying Trained Model on South American Extent {.tabset}

#### MaxEnt Model for South American Extent

```{r predict_south_america}
# Apply the Asia model to South America.

source("scripts/utils/read_in_predictor_data_south_america.R")

south_america_preds = terra::predict(south_america_r, asia_best_model, type = "response", na.rm=T)

ggplot() + 
  tidyterra::geom_spatraster(data = south_america_preds) + 
  geom_point(data = south_american_occs, aes(x=x, y=y)) + 
  scale_fill_distiller(palette = 'Spectral') + 
  labs(fill = 'Suitable\nHabitat')
```

#### MaxEnt Model Transformed: "Habitat-or-Not" Threshold Map

```{r maxent_plot_south_america}
# Apply the minimum suitable habitat for which we found occurrences in asia,
# to calculate a habitat-or-not for South America.

south_america_preds_hab_or_not = invisible(south_america_preds > asia_min_suitability_with_occ)

ggplot() + 
  tidyterra::geom_spatraster(data = south_america_preds_hab_or_not) + 
  geom_point(data = south_american_occs, aes(x=x, y=y)) + 
  scale_fill_manual(values = c("FALSE" = "lightgreen", "TRUE" = "purple"), na.value = 'transparent') +
  labs(fill = 'Suitable\nHabitat')
```

### Applying Trained Model on North American Extent {.tabset}

#### MaxEnt Model for North American Extent

```{r predict_north_america_and_plot}
# ---------------------------------------
# STEP 3: APPLY MODEL TO NORTH AMERICA
# ---------------------------------------
source("scripts/utils/read_in_predictor_data_north_america.R")
# Apply the Asia model to North America.
north_america_preds = terra::predict(north_america_r, asia_best_model, type = "response", na.rm=T)

ggplot() + 
  tidyterra::geom_spatraster(data = north_america_preds) + 
  geom_point(data = north_american_occs, aes(x=x, y=y)) + 
  scale_fill_distiller(palette = 'Spectral') + 
  labs(fill = 'Suitable\nHabitat')
```

#### MaxEnt Model Transformed: "Habitat-or-Not" Threshold Map

```{r habitat_north_america}
# Apply the minimum suitable habitat for which we found occurrences in asia,
# to calculate a habitat-or-not for north America.

north_america_preds_hab_or_not = invisible(north_america_preds > asia_min_suitability_with_occ)

ggplot() + 
  tidyterra::geom_spatraster(data = north_america_preds_hab_or_not) + 
  geom_point(data = north_american_occs, aes(x=x, y=y)) + 
  scale_fill_manual(values = c("FALSE" = "lightgreen", "TRUE" = "purple"), na.value = 'transparent') +
  labs(fill = 'Suitable\nHabitat')

```

### Applying Trained Model on just British Columbia {.tabset}

#### MaxEnt Model for BC Extent

```{r predict_bc_and_plot}
bc_preds = terra::mask(terra::crop(north_america_preds, bc), bc)

ggplot() + 
  tidyterra::geom_spatraster(data = bc_preds) + 
  scale_fill_distiller(palette = 'Spectral') + 
  labs(fill = 'Suitable\nHabitat')
```

#### MaxEnt Model Transformed: "Habitat-or-Not" Threshold Map

```{r bc_habitat}
# Apply the minimum suitable habitat for which we found occurrences in asia,
# to calculate a habitat-or-not for north America.

bc_preds_hab_or_not = invisible(bc_preds > asia_min_suitability_with_occ)

ggplot() + 
  tidyterra::geom_spatraster(data = bc_preds_hab_or_not) + 
  scale_fill_manual(values = c("FALSE" = "lightgreen", "TRUE" = "purple"), na.value = 'transparent') +
  labs(fill = 'Suitable\nHabitat')

```